Modelling state-switching functional biologging data with hidden Markov models

### 2020-11-23 Report/Sections/simulation.tex
Address comments (still need to clean figures)
Updates from Overleaf

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 25s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$ as defined in Equation (\ref{eqn:z}). Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
    %
    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
    %
    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows one realization of $Z^*$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

@@

The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model. The average dive decoding accuracy is above 0.9 while the average subdive decoding accuracy is $1-10^{-12}$. All parameter estimates of mean values and fine-scale probability transition matrices, $\hat \mu$ and $\hat \Gamma^*$, respectively, have statistically insignificant biases. In addition, biases for $\hat \sigma$, $\hat \phi$, and $\hat \Gamma$ are as severe or less severe than the other three models. The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to slightly underestimate the empirical standard errors. See Tables 6 through 9 of the supplementary material for detailed results.

>>>>

We generate 500 independent data sets from the CarHHMM-DFT. Each data set consists of a sequence of 100 curves which we call a sequence of killer whale dives. Each dive can be one of $N=2$ dive types based on a Markov chain with probability transition matrix
%
$$\Gamma = \begin{pmatrix} 0.5 & 0.5 \\ 0.5 & 0.5 \end{pmatrix}.$$
%
Dive duration is Gamma distributed, and the coarse-scale emission parameters $\theta$ are $\mu^{(1)} = 20s$, $\sigma^{(1)} = 5s$, $\mu^{(2)} = 80s$, and $\sigma^{(2)} = 50s$. After generating the dive durations for all 100 dives in a data set, dive $t$ is broken into a sequence of $\tilde T^*_t = \lfloor Y_t/2 \rfloor$ two-second windows, where the last $Y_t - 2 \tilde T^*_t$ seconds of each dive are ignored. Each two-second segment is assigned one of $N^*=2$ behaviours according to a fine-scale Markov chain $\tilde X^*_t \equiv \left\{X^*_{t,1}, \ldots \right\}$ with probability transition matrices
%
$$\Gamma^{*(1)} = \begin{pmatrix} 0.5 & 0.5 \\ 0.1 & 0.9 \end{pmatrix} \quad \text{ and } \quad \Gamma^{*(2)} = \begin{pmatrix} 0.8 & 0.2 \\ 0.3 & 0.7 \end{pmatrix}$$ 
%
for dive types 1 and 2, respectively.
Instead of generating the raw observations $Y^*_{t,t^*}$, we directly simulate the fine-scale transformed observations $\Z_{t,\tilde t^*} = \left\{\Zone_{t,\tilde t^*}, \Ztwo_{t,\tilde t^*}\right\}$ as defined in Equation (\ref{eqn:z}). Recall that $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ are independent when conditioned on all dive types and subdive states. Each observation of acceleration $\Zone_{t,\tilde t^*}$ is Normally-distributed with explicit auto-correlation while each observation of wiggliness $\Ztwo_{t,\tilde t^*}$ is Gamma-distributed without explicit auto-correlation. In particular, we set the distributions of $\Zone_{t,\tilde t^*}$ and $\Ztwo_{t,\tilde t^*}$ with the following parameters:
%
\begin{gather*}
    \mu_A^{*(\cdot,1)} = 0.0 s, \enspace \sigma_A^{*(\cdot,1)} = 0.05s, \enspace \phi_A^{*(\cdot,1)} = 0.99, \\
    %
    \mu_A^{*(\cdot,2)} = 0.0 s, \enspace \sigma_A^{*(\cdot,2)} = 0.1s, \enspace \phi_A^{*(\cdot,2)} = 0.95, \\
    %
    \mu_W^{*(\cdot,1)} = 10.10, \quad \sigma_W^{*(\cdot,1)} = 3.18, \\
    %
    \mu_W^{*(\cdot,2)} = 305.94, \quad \sigma_W^{*(\cdot,2)} = 17.49.
\end{gather*}
%
It is not possible to uniquely reconstruct the raw accelerometer data $Y^*$ from $Z^*$ alone, but we describe one possible mapping from $Z^*$ to $Y^*$ in the appendix. Figure \ref{fig:sim_data} shows one realization of $Z^*$ for five dives of one simulated data set along with the the corresponding reconstructed realization of $Y^*$.

@@

The full CarHHMM-DFT is the best performing model of the four candidates since it is the generating model. The average dive decoding accuracy is greater than 0.9 while the average subdive decoding accuracy is $1-10^{-12}$. All parameter estimates of mean values and fine-scale probability transition matrices, $\hat \mu$ and $\hat \Gamma^*$, respectively, have statistically insignificant biases. In addition, biases for $\hat \sigma$, $\hat \phi$, and $\hat \Gamma$ are as severe or less severe than the other three models. The empirical standard errors of all parameter estimates ($\hat \theta$, $\hat \Gamma$, $\hat \theta^*$, $\hat \Gamma^*$) are well-approximated by the inverse of the observed Fisher information matrix, although the estimated standard errors tend to slightly underestimate the empirical standard errors. See Tables 6 through 9 of the supplementary material for detailed results.
